Interactive web-based flood modeling at country wide scale and planter size resolution.

### 2014-05-16 review/rebuttal.txt

## Reviewer 1
> Review:	Check for comma rules on all enumerations througout the paper,
> put a comma before the last "and".

serial commas used, it's/its fixed

> Page 3, para 3:
> "determines", "one's"

determines-> determine
one's ones used correctly

> page 5, para 4:
> put units after "12" and "50" (fps ?)
Correct latex macro \si -> \SI used, fixes units throughout the paper

> page 5, para 5:
> "Malen nach Zahlen"

Changed \emph to ``''

## Reviewer 2
> Review:	This paper describes the 3Di platform, a new overland flooding and hydrological model. The paper primarily discusses the model's software structure (authors call this "numerical model), architecture, and visualization techniques. The paper is divided into 3 primary sections corresponding to these categories.
> 1. Typo - Section 2, Paragraph 7:
> "To represent water systems in civilized regions the 2D grid was extended <<insert:with>> 1D channels"

Changed

> 2. Typo - S2, para 8:
> "the subgrid concept is also applied <<insert:to>> the hydrological processes in the subsoil"

Changed

> 3. Typo - S2, para 9: oxford comma - drainage resistance, and maximum infiltration

Serial commas applied consistently

> 4. Missing discussion - S3, Para 1:
> The author's answers to "how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling", is to "use the web" based on the corresponding text. How is the web based approach adequate for tablets and mobile phones? There is no discussion regarding the use of these devices.

Added in system description and in discussion.

> 5. Typo - S3, para 4: "If no indication of progress is given for longer than two second<<insert:s>> users will start reloading"

Units are fixed throughout the paper

> 6. Bad reference - The reference to Baart 2014 is incomplete. There is no document, article or wbesite described. How does one access this content?
DOI added to sourcecode of the python implementation.

> 7. Confusion - What is the purpose of the "bytes" following the JSON in a model message? This is not described at all.
Bytes replaced by serialized array.

> 8. Typo - S3, para 5 "The advantage of using the MMI, 0MQ combination is that it has flexible metadata, does not copy data, <<insert:and>> is very fast.

Changed.

> 9. Typo - S3, para 6 "User interactions are directly <<change to:sent>> through websockets
Changed.

> 10. Typo - S3, para 6, "After a model timestep <<insert:,>> the chain is inverted.

Changed

> 11. Typo - S3, para 6, "This fast feedback loop <<change to:supports user interactivity with the model.>>

Changed

> 12. Revise - S3, para 7, "The social aspect of modeling is another aspect"

The social aspect is another source of interesting problems.

> 13. Clarify - S4, para 2, What are the units for the numbers in this paragraph? Are the milliseconds? seconds? The reader is left to infer the units. I guess they are probably milliseconds.

Units fixed throughout the paper

> 14. Explain - "The first is to reduce the timestemp so that it computes in less that 50."
> Are you referring to the modeling timestep? Please clatify.

Changed to modeling timestep

> 15. S3, para 4 Define what a voxel is.

Defined as a picture element (pixel) with volume.

> 16. Typo - S3, para 6 "One can argue whether to show these 'true' yet confusing results or to process the information so that <<insert:it>> is perceived as intended."

Changed

> 17. Figure 4 & 5 captions - add the interpolation method type, e.g. barycentric...

Changed

> 18. Typo - S3, para 7, Revise third sentence to "Is also allows water level results to be interpolated."

Changed to It...

> 19. Typo - S3, para 7, "from the cell centers <<delete:and>> is linearly interpolated

Changed

> 20. Typo - S3, para 7, "uses a weighted average <<change:to>> determine water levels."

Changed

> 21. Revise - S3, para 8, "<<delete:The>> Triangulation with the barycentric <<moved:linear>> interpolation also allow us to

The triangular mesh can be reused to filter levee artefacts. The triangulation can reconstruct levees with use of additional levee geometries.

> 22. The conclusion is weak.

> 23. Sections 2 and 3 seem to overlap. Both contain various aspects regaarding the software framework and architecture. The paper has not been cleanly organized.

> 24. There is no discussion regarding the accuracy of the flood model itself. The paper primarily presents the software structure, features, and visualization capabilities.

> 25. The authors briefly mention how 3Di supports collaboration. It would be nice to see more discussion.

> 26. This paper appears to discuss both the 3Di modeling platform/framework and a new flooding/hydrological model. Does the flood model have a name?

> 27. Typo: Section 1, After subtitles add ":"'s, for example "Mono-disciplinary versus integrated <<insert ":">>"

Changed

> 28. The discussion and relevance to cloud based modeling is very limited in this paper. There is only a brief mention of "IaaS" and "Amazon". It is unclear what has been tested and run using cloud computing techniques.

> 29. There is no discussion from the abstract on how or if 3Di achieves "100 times faster than the previous approach". What is the previous approach? What performance evaluations have been conducted?

### 2014-05-16 article.tex
textual changes, spelling

These challenges arise from the movement towards more modern working methods and the ever increasing data volumes. The working methods are changing on different aspects.
\begin{itemize}
\item \emph{Mono-disciplinary versus integrated} Models used to be created by one or few people working in the same discipline. The challenge now is to model the integrated environment, which can best be approached as a community \citep{Voinov2010} of different disciplines.
\item \emph{Confirmatory versus exploratory} From assuming that the model is valid and can be used for hypothesis testing to realizing it's a made-up simplification of reality \citep{Oreskes1994} that can be used to explore possibilities.
\item \emph{Passive versus interactive} Moving from the pre-processing, wait a few hours, post-processing cycle towards a millisecond feedback loop with high resolution and connected physical processes. The challenge here is to move from the scale of a square kilometer \citep[for example][]{Losasso2008} to the scale of small country.
\item \emph{Solitary versus social} The solitary modeler creating static pictures for a 600 page report describing all the different scenarios to a group of stakeholders in control \footnote{preferably with guidance from domain experts} of a model running simultaneous on their mobile devices and a big screen in the control room.
\end{itemize}

@@

The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}. To represent water systems in civilized regions the 2D grid was extended 1D channels and controllable structures (for example pumps, orifices, weirs). This allows for more realistic and interactive overland flooding simulation.

The power of the subgrid concept is also applied the hydrological processes in the subsoil. An important factor in potential flooding is the storage of water in the subsoil, or the lack thereof. The change in surface water volume is driven by several familiar processes:
\[
\Delta V_\mathrm{surface\ water} = (Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evaporation} + Q_\mathrm{drainage})\cdot\Delta t
\]

@@

The end-users, interact with the web-based \ac{GUI}. The design philosophy is minimalistic, allowing little room for user mistakes. For example discharge locations only have two options (normal and extreme discharge) (see \autoref{fig:twod}). Sessions with groups of end-users resulted in this design that provides freedom to explore without the insinuation that the model results are ``true''.

@@

To communicate the model results the \ac{MMI} protocol is used \citep{Baart2014a}. This protocol is suitable for sending model variables over the internet. A model message consists of a metadata header in \ac{JSON} format, optionally followed by bytes. Listing \ref{lst:mmi} shows an example of an \ac{MMI} message. In this project the \ac{0MQ} protocol is used as a the underlying transport protocol. The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data is very fast.

\begin{lstlisting}[caption=MMI message,label=lst:mmi]
# The first part of the message contains metadata
{
  "name": "waterlevel",
  "shape": [3,3],
  "dtype": "float64",
  "attributes": {
       "standard_name": "sea_surface_altiude",
       "units": "m"
  }
}
# The second part of the message contains bytes
\x81\xb3b\xa0\xe8 ... \x01\x90\xe5\x84\xb5\xb0
\end{lstlisting}

Interaction is another challenge where timing is crucial. When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice, the user expects a response within a second. This short feedback loop is achieved using several techniques. User interactions are directly send through websockets \citep{Hickson2012} to the model controller. These interactions are cached and pushed into the model on every timestep. These timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common. After a model timestep the chain is inverted. The model pushes it's arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results. This fast feedback loop gives a good sense of interactivity.

The social aspect of modeling is another aspect where interesting problems occur. Several group sessions with different settings have been tried. The current setup is that the model is run in a ``virtual room'', like a chat room. Multiple people can join the model run. One person is in control and others can view and zoom in on the same model run. Other users can request control of the model. The sessions worked best when a topic expert, not necessary a modeler, was present in the session.

@@

First a note about the performance. Humans will detect images at a rate of about \SI{12}{\per\second} as continuous \citep{Landis1954}. To enable the model results to be perceived as continuous, taking into account some random network delay, the aim is to generate a map within \SI{50}{\milli\second}. Generating results at this framerate can be done using two approaches. The first is to reduce the timestep so that it computes in less than \SI{50}{\milli\second}. This is the only way to get true model results at this framerate. An alternative is to interpolate in time and provide the insinuation of continuous information. We use both approaches here. We reduce the calculation time so that interaction is fluent and we can pass the flicker fusion threshold. If one wants to use longer time steps we also visualize at a high framerate. An example of this is the visualization of river flow using dots that are animated with \ac{CSS} animations as a function of the current flow velocity (see \autoref{fig:oned}).

@@

The variety of end-users exposes the numerical model to non-technical users with little appreciation of implemented numerical schemes. Some artifacts from the numerical schemes result in confusion. The world in the subgrid model is internally represented in voxels, not unlike the computer game Minecraft. This results in visual artifacts, as represented in \autoref{fig:arta}. Another type of artifact occurs when a levee crosses a calculation cell both sides of the cell will contain water.

One can argue whether to show these ``true'' yet confusing results or to process the information so that is perceived as intended. In the \threedi platform the default visualization hides these artifacts by filtering and smoothing the output. Hiding the square world is implemented using different techniques.

\begin{figure}[htpb]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{arta}
    \caption{Uninterpolated subgrid water level}
    \label{fig:arta}
  \end{subfigure}
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{artb}
    \caption{Interpolated subgrid water level}
    \label{fig:artb}
  \end{subfigure}
  \label{fig:artifacts}
  \caption{Interpolation method}
\end{figure}

A barycentric linear interpolation technique is used. This is one of the few techniques that allow for the interpolation within the rendering time constraint. It also allows to interpolate water level results on an arbitrary grid. The quad tree grid is triangulated using Qhull \citep{Barber1996}.  Water level information, from the cell centers and is linearly interpolated to the underlying subgrid. The interpolation uses a weighted average the determine water levels. The triangular mesh can be reused.   \autoref{fig:artb} shows the results of this technique .

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{levees1}
    \caption{Redistribution of cell sections based on levee positioning.}
    \label{fig:levee1}
  \end{subfigure}
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{levees2}
    \caption{Interpolated subgrid water level with levee constrained}
    \label{fig:levee2}
  \end{subfigure}
  \label{fig:levees}

@@

The triangulation within the linear barycentric interpolation also allows us to filter levee artefacts of the numerical model. The calculation grid incorporates the function of a levee in the numerical schemes. However it disregards the actual position of such a levee within a calculation cell. The triangulation can reconstruct levees with use of additional levee geometries. A triangle is one of the simplest geometrical forms and can be used reconstruct any arbitrary shape \citep{Welch1994}. To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (\autoref{fig:levee1}). For the intersected calculation cells the influential water level points are determined and coupled to the relevant calculation areas. The arrows in \autoref{fig:levee1} indicate which water level is relevant for the dike intersected cell areas. When all waterlevel points are allocated to the dike, the triangulation can be executed with the incorporation of the levee geometries. The newly created triangular mesh now also follows the added levee geometry. Successively the interpolation is executed, resulting in water being retained behind the dike within one calculation cell. The interpolated result is shown in \autoref{fig:levee2}.

@@

The major advancement here is to redesign the numerical engine into a component that allows for external control and introspection. By combining this numerical model with the technical advancements described in this paper, we have enabled interactive web-based modeling at a high resolution and wide coverage.

>>>>

These challenges arise from the movement towards more modern working methods and the ever increasing data volumes. The working methods are changing on different aspects.
\begin{itemize}
\item \emph{Mono-disciplinary versus integrated:} Models used to be created by one or few people working in the same discipline. The challenge now is to model the integrated environment, which can best be approached as a community \citep{Voinov2010} of different disciplines.
\item \emph{Confirmatory versus exploratory:} From assuming that the model is valid and can be used for hypothesis testing to realizing it's a made-up simplification of reality \citep{Oreskes1994} that can be used to explore possibilities.
\item \emph{Passive versus interactive:} Moving from the pre-processing, wait a few hours, post-processing cycle towards a millisecond feedback loop with high resolution and connected physical processes. The challenge here is to move from the scale of a square kilometer \citep[for example][]{Losasso2008} to the scale of small country.
\item \emph{Solitary versus social:} The solitary modeler creating static pictures for a 600 page report describing all the different scenarios to a group of stakeholders in control \footnote{preferably with guidance from domain experts} of a model running simultaneous on their mobile devices and a big screen in the control room.
\end{itemize}

@@

The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}. To represent water systems in civilized regions the 2D grid was extended with 1D channels and controllable structures (for example pumps, orifices, weirs). This allows for more realistic and interactive overland flooding simulation.

The power of the subgrid concept is also applied to the hydrological processes in the subsoil. An important factor in potential flooding is the storage of water in the subsoil, or the lack thereof. The change in surface water volume is driven by several familiar processes:
\[
\Delta V_\mathrm{surface\ water} = (Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evaporation} + Q_\mathrm{drainage})\cdot\Delta t
\]

@@

The end-users, interact with the web-based \ac{GUI}. The design philosophy is minimalistic, allowing little room for user mistakes. For example discharge locations only have two options (normal and extreme discharge) (see \autoref{fig:twod}). Sessions with groups of end-users resulted in this design that provides freedom to explore without the insinuation that the model results are ``true''. The application of \ac{CSS} and \ac{SVG} allows the interface to scale between different screen resolutions so that it can be used on different devices (phones, tablets, browsers, map table).

@@

To communicate the model results the \ac{MMI} protocol is used \citep{Baart2014a}. This protocol is suitable for sending model variables over the internet. A model message consists of a metadata header in \ac{JSON} format, optionally followed by the serialized array. Listing \ref{lst:mmi} shows an example of an \ac{MMI} message. In this project the \ac{0MQ} protocol is used as a the underlying transport protocol. The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data, and is very fast.

\begin{lstlisting}[caption=MMI message,label=lst:mmi]
# The first part of the message contains metadata
{
  "name": "waterlevel",
  "shape": [3,3],
  "dtype": "float64",
  "attributes": {
       "standard_name": "sea_surface_altiude",
       "units": "m"
  }
}
# The second part of the message contains a serialized array
\x81\xb3b\xa0\xe8 ... \x01\x90\xe5\x84\xb5\xb0
\end{lstlisting}

Interaction is another challenge where timing is crucial. When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice, the user expects a response within a second. This short feedback loop is achieved using several techniques. User interactions are directly sent through websockets \citep{Hickson2012} to the model controller. These interactions are cached and pushed into the model on every timestep. These timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common. After a model timestep, the chain is inverted. The model pushes its arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results. This fast feedback loop supports user interactivity with the model.

The social aspect is another source of interesting problems. Several group sessions with different settings have been tried. The current setup is that the model is run in a ``virtual room'', like a chat room. Multiple people can join the model run. One person is in control and others can view and zoom in on the same model run. Other users can request control of the model. The sessions worked best when a topic expert, not necessary a modeler, was present in the session.

@@

First a note about the performance. Humans will detect images at a rate of about \SI{12}{\per\second} as continuous \citep{Landis1954}. To enable the model results to be perceived as continuous, taking into account some random network delay, the aim is to generate a map within \SI{50}{\milli\second}.
Generating results at this framerate can be done using two approaches. The first is to reduce the modeling timestep so that it computes in less than \SI{50}{\milli\second}. This is the only way to get true model results at this framerate.
An alternative is to interpolate in time and provide the insinuation of continuous information.
We use both approaches here. We reduce the calculation time so that interaction is fluent and we can pass the flicker fusion threshold. If one wants to use longer time steps we also visualize at a high framerate. An example of this is the visualization of river flow using dots that are animated with \ac{CSS} animations as a function of the current flow velocity (see \autoref{fig:oned}).

@@

The variety of end-users exposes the numerical model to non-technical users with little appreciation of implemented numerical schemes. Some artifacts from the numerical schemes result in confusion. The world in the subgrid model is internally represented in voxels (a picture element (pixel) with volume), not unlike the computer game Minecraft. This results in visual artifacts, as represented in \autoref{fig:arta}. Another type of artifact occurs when a levee crosses a calculation cell both sides of the cell will contain water.

One can argue whether to show these ``true'' yet confusing results or to process the information so that it is perceived as intended. In the \threedi platform the default visualization hides these artifacts by filtering and smoothing the output. Hiding the square world is implemented using different techniques.

\begin{figure}[htpb]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{arta}
    \caption{Uninterpolated subgrid water level}
    \label{fig:arta}
  \end{subfigure}
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{artb}
    \caption{Interpolated subgrid water level}
    \label{fig:artb}
  \end{subfigure}
  \label{fig:artifacts}
  \caption{Linear barycentric interpolation method}
\end{figure}

A barycentric linear interpolation technique is used. This is one of the few techniques that allow for the interpolation within the rendering time constraint. It also allows water level results to be interpolated. The quad tree grid is triangulated using Qhull \citep{Barber1996}.  Water level information, from the cell centers is linearly interpolated to the underlying subgrid. The interpolation uses a weighted average to determine water levels. \autoref{fig:artb} shows the results of this technique .

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{levees1}
    \caption{Redistribution of cell sections based on levee positioning.}
    \label{fig:levee1}
  \end{subfigure}
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=1\textwidth]{levees2}
    \caption{Interpolated subgrid water level with levee constraint}
    \label{fig:levee2}
  \end{subfigure}
  \label{fig:levees}

@@

The triangular mesh can be reused to filter levee artefacts. The triangulation can reconstruct levees with use of additional levee geometries. A triangle is the simplest geometrical form and can be used reconstruct any arbitrary shape \citep{Welch1994}. To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (\autoref{fig:levee1}). For the intersected calculation cells the influential water level points are determined and coupled to the relevant calculation areas. The arrows in \autoref{fig:levee1} indicate which water level is relevant for the dike intersected cell areas. When all waterlevel points are allocated to the dike, the triangulation can be executed with the incorporation of the levee geometries. The newly created triangular mesh now also follows the added levee geometry. Successively the interpolation is executed, resulting in water being retained behind the dike within one calculation cell. The interpolated result is shown in \autoref{fig:levee2}.

@@

The major advancement here is to redesign the numerical engine into a component that allows for external control and introspection. By combining this numerical model with the technical advancements described in this paper, we have enabled interactive web-based modeling at a high resolution and wide coverage. The use of modern web technologies allows the same interface to be used on a browser, a phone and on a map table.
